---
title: "ALCL TME signature among PTCL"
date: '`r Sys.Date()`'
output:
    html_notebook:
    fig_height: 12
    fig_width: 10
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}

library(devtools)
library(ggplot2)
library(RColorBrewer)
library(tidyr)
library(viridis)
library(pheatmap)
library(ConsensusClusterPlus)
library(ComplexHeatmap)
library(circlize)
library(ggpubr)
```


# ALCL (ALCL+ and ALCL-) VS other PTCL (oPCTL : PTCL nos, AITL, PTCL nos/AITL)

## 1. Functions & Metadata

```{r, warning = FALSE}

# useful functions

hcopt <- function(d, HC=NULL, method = "ward.D", members = NULL){
  require("cba")
  if ( is.null(HC) ) {
    HC <- hclust(d,method=method,members=members)
  }
  #optimal leaf ordering
  ORD <- cba::order.optimal(d,merge=HC$merge)
  HC$merge <- ORD$merge
  HC$order <- ORD$order
  HC
}

colGradient <- function( cols, length, cmax=255 )
{
  ## e.g., to create a white-to-red gradient with 10 levels
  ##   colGradient(cols=c('white','red'),length=10)
  ## or, to create a blue-to-white-to-red gradients with 9 colors
  ##   colGradient(cols=c('blue','white','red'),length=9)
  ramp <- grDevices::colorRamp(cols)
  rgb( ramp(seq(0,1,length=length)), max=cmax )
}

# metadata

metadata <- read.table(file = "./cel-affymetrix_ALCL_oPTCL_metadata.csv", header = TRUE)
metadata <- metadata[order(metadata$name),]
head(metadata)
dim(metadata)
```

## 2. CIBERSORT (online) with LM22 matrix + 2 stromal cell types from MCP-Counter on Affy expression data

```{r, warning = FALSE}

TME_estim <- as.matrix(read.table(file = "./CIBERSORTx-LM22_and_MCP-counter-stroma_Affy-ALCL-oPTCL_patients.csv", header = TRUE, row.names = 1))
TME_estim <- TME_estim[which(rownames(TME_estim)%in%metadata$name),]
dim(TME_estim)
head(TME_estim[,1:5])
```

## 3. Graphical overviews 

### 3.1. Graphical overview of ALCL VS oPTCL (Heatmap Fig. 1)

```{r, warning = FALSE, fig.width=12, fig.height=14}

# create a dataframe for ggplot2 figures
d <- as.data.frame(cbind(sample=rownames(TME_estim),TME_estim)) %>% 
gather(key = "cell_type", value = "score", 2:25, na.rm = TRUE, factor_key=TRUE)
d$score <- as.numeric(d$score)
d_metadat <- cbind(cbind(patho=rep(metadata$patho,24),specific_patho=rep(metadata$specific_patho,24)),d)
d_metadat <- d_metadat[order(d_metadat$patho),]
d_metadat$sample <- factor(d_metadat$sample, levels = unique(d_metadat$sample))
# scaled data (Z-score) and transposed
data <- t(scale(as.matrix(TME_estim[,1:24])))
dim(data)

# clustering / k choice
CCout <- ConsensusClusterPlus(data,maxK=3,reps=500,pItem=0.9,pFeature=1,
                              innerLinkage="ward.D", finalLinkage="ward.D",
                              title="cc",clusterAlg="hc",distance="euclidean", plot=NULL)
summary(CCout)
nc <- 3 # variable for the following commands

# Heatmap parameters
annot1 <- data.frame("PTCL_type"=as.factor(metadata[,c("patho")]),"PTCL_subtype"=as.factor(metadata[,c("specific_patho")]),"cluster"=CCout[[nc]]$consensusClass)
annot1$cluster <- as.factor(annot1$cluster)
rownames(annot1) <- metadata$name
annotCol <- list(PTCL_type = c("ALCL" = "#9b4ddb", "oPTCL" = "#37a695"), PTCL_subtype = c("AITL" = "#96b0a1", "PTCL_nos" = "#387d4d", "PTCL_nos/AITL" = "#e6e1b5", "ALCL-" = "#dfcfe6", "ALCL+" = "#805194"),cluster = c("1" = "#d9969c", "2" = "#8dc9af", "3" = "#6e7fb5"))
## use cluster tree from consensus clustering for column ordering in heatmap
clust.col <- CCout[[nc]]$consensusTree
## determine row ordering based on de-novo clustering
clust.row <- hcopt(as.dist(1-cor(t(data))),method="ward.D")

# Color range for the heatmap
col_fun <- colorRamp2(c(-5, 0, 5), c("blue", "white", "red")) # Z-score
# HeatmapAnnotation object for up annotations
ha <- HeatmapAnnotation(df = annot1, col = annotCol,
  annotation_name_side = "left",   # position du nom de l'annotation
  annotation_legend_param = list(title_position = "topcenter",  # légende au-dessus et centrée
    nrow = 1                      # online legend
  )
)
# Heatmap
ht <- Heatmap(
  data, name = "scaled score", # nom du colorbar
  col = col_fun, top_annotation = ha,
  cluster_rows = clust.row, cluster_columns = clust.col,
  show_row_names = TRUE, show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 12),   # taille du texte pour les lignes
  column_names_gp = gpar(fontsize = 10), # taille du texte pour les colonnes (au cas où)
  column_split = 3,                     # couper les colonnes en 3 groupes
  heatmap_legend_param = list(title_position = "topcenter",
    legend_direction = "horizontal"    # orientation horizontale de la légende
  )
)
# 4. Dessiner avec la légende au-dessus
pdf("heatmap_global.pdf", width = 10, height = 8)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()
```

### 3.2. Graphical overview of ALCL VS AITL only (Heatmap Supplementary Fig. 1)

```{r, warning = FALSE, fig.width=12, fig.height=14}

metadata_short <- metadata[which(metadata$specific_patho=="AITL" | metadata$patho=="ALCL"),]
TME_estim_short <- TME_estim[which(rownames(TME_estim)%in%metadata_short$name),]

# create a dataframe for ggplot2 figures
d <- as.data.frame(cbind(sample=rownames(TME_estim_short),TME_estim_short)) %>% 
gather(key = "cell_type", value = "score", 2:25, na.rm = TRUE, factor_key=TRUE)
d$score <- as.numeric(d$score)
d_metadat <- cbind(cbind(patho=rep(metadata_short$patho,24),specific_patho=rep(metadata_short$specific_patho,24)),d)
d_metadat <- d_metadat[order(d_metadat$patho),]
d_metadat$sample <- factor(d_metadat$sample, levels = unique(d_metadat$sample))
# scaled data (Z-score) and transposed
data <- t(scale(as.matrix(TME_estim_short[,1:24])))
dim(data)

# clustering / k choice
CCout <- ConsensusClusterPlus(data,maxK=3,reps=500,pItem=0.9,pFeature=1,
                              innerLinkage="ward.D", finalLinkage="ward.D",
                              title="cc",clusterAlg="hc",distance="euclidean", plot=NULL)
summary(CCout)
nc <- 2 # variable for the following commands

# side and heatmap color - function of ALCL (ALK+ et -) vs other T
my_group <- as.numeric(as.factor(metadata_short$patho))
colSide <- brewer.pal(5, "Set2")[my_group]
colMain <- colorRampPalette(brewer.pal(8, "YlGnBu"))(25)
# heatmap
heatmap(TME_estim_short[,1:24], RowSideColors=colSide, col=colMain,margins = c(11,4))
legend(x = "right", inset=0.01, legend = c("low", "medium", "high"), cex = 0.8, fill = colorRampPalette(brewer.pal(8, "YlGnBu"))(3))

# Heatmap parameters
annot1 <- data.frame("PTCL_type"=as.factor(metadata_short[,c("patho")]),"PTCL_subtype"=as.factor(metadata_short[,c("specific_patho")]),"cluster"=CCout[[nc]]$consensusClass)
annot1$cluster <- as.factor(annot1$cluster)
rownames(annot1) <- metadata_short$name
annotCol <- list(PTCL_type = c("ALCL" = "#9b4ddb", "oPTCL" = "#37a695"), PTCL_subtype = c("AITL" = "#96b0a1", "ALCL-" = "#dfcfe6", "ALCL+" = "#805194"),cluster = c("1" = "#d9969c", "2" = "#8dc9af"))
## use cluster tree from consensus clustering for column ordering in heatmap
clust.col <- CCout[[nc]]$consensusTree
## determine row ordering based on de-novo clustering
clust.row <- hcopt(as.dist(1-cor(t(data))),method="ward.D")

# Color range for the heatmap
col_fun <- colorRamp2(c(-5, 0, 5), c("blue", "white", "red")) # Z-score
# HeatmapAnnotation object for up annotations
ha <- HeatmapAnnotation(df = annot1, col = annotCol,
  annotation_name_side = "left",   # position du nom de l'annotation
  annotation_legend_param = list(title_position = "topcenter",  # légende au-dessus et centrée
    nrow = 1                      # online legend
  )
)
# Heatmap
ht <- Heatmap(
  data, name = "scaled score", # nom du colorbar
  col = col_fun, top_annotation = ha,
  cluster_rows = clust.row, cluster_columns = clust.col,
  show_row_names = TRUE, show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 12),   # taille du texte pour les lignes
  column_names_gp = gpar(fontsize = 10), # taille du texte pour les colonnes (au cas où)
  column_split = 3,                     # couper les colonnes en 3 groupes
  heatmap_legend_param = list(title_position = "topcenter",
    legend_direction = "horizontal"    # orientation horizontale de la légende
  )
)
# 4. Dessiner avec la légende au-dessus
pdf("heatmap_ALCL_AITL.pdf", width = 10, height = 8)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()
```

## 4. Compute p-values for each cell type comparison (boxplots)

### 4.1. Graphical overview of ALCL VS oPTCL for each cell type (boxplots not included)

```{r, warning = FALSE}

d <- as.data.frame(cbind(sample=rownames(TME_estim),TME_estim)) %>% 
gather(key = "cell_type", value = "score", 2:25, na.rm = TRUE, factor_key=TRUE)
d$score <- as.numeric(d$score)
d_metadat <- cbind(cbind(patho=rep(metadata$patho,24),specific_patho=rep(metadata$specific_patho,24)),d)
d_metadat <- d_metadat[order(d_metadat$patho),]
d_metadat$sample <- factor(d_metadat$sample, levels = unique(d_metadat$sample))

# compute p-values between ALCL and oPTCL groups for the different cell types
perform_wilcox_test <- function(d, metadata) {
  wilcox_pval <- c()
  cells <- unique(d$cell_type)
  metadata <- metadata
  for (c in cells) {
    mat <- cbind(patho = metadata$patho, d[which(d$cell_type == c), ])
    if (length(unique(mat$patho)) > 1) {  # check that we have 2 groups
      pval <- wilcox.test(score ~ patho, data = mat)$p.value
    } else {
      pval <- NA
    }
    wilcox_pval[c] <- pval
  } # p-values correction
  p_corrected <- p.adjust(wilcox_pval, method = "hochberg", n = length(wilcox_pval))
  return(p_corrected)
}

p <- perform_wilcox_test(d, metadata)
# significative
p[which(p<0.00001)] #1^10-5

# Boxplots examples
type="NK_cells_resting"
my_comparisons <- list( c("ALCL", "oPTCL"))
d_short <- d_metadat[which(d_metadat$cell_type==type),]
ggplot(d_short, aes(x=patho, y=score, fill=patho)) +  labs(title=type)+
geom_boxplot(outlier.colour="darkgrey",outlier.size=2) + scale_fill_manual(name = "PTCL", labels = c("ALCL", "oPTCL"), values=c("#9b4ddb", "#37a695")) + theme_bw(base_size = 20) + xlab("") + stat_compare_means(comparisons = my_comparisons, size=4)+scale_x_discrete(labels=c("ALCL","oPTCL")) + theme(legend.position="none")
type="T_cells_follicular_helper"
d_short <- d_metadat[which(d_metadat$cell_type==type),]
ggplot(d_short, aes(x=patho, y=score, fill=patho)) +  labs(title=type)+
geom_boxplot(outlier.colour="darkgrey",outlier.size=2) + scale_fill_manual(name = "PTCL", labels = c("ALCL", "oPTCL"), values=c("#9b4ddb", "#37a695")) + theme_bw(base_size = 20) + xlab("") + stat_compare_means(comparisons = my_comparisons, size=4)+scale_x_discrete(labels=c("ALCL","oPTCL")) + theme(legend.position="none")
type="Fibroblasts"
d_short <- d_metadat[which(d_metadat$cell_type==type),]
ggplot(d_short, aes(x=patho, y=score, fill=patho)) +  labs(title=type)+
geom_boxplot(outlier.colour="darkgrey",outlier.size=2) + scale_fill_manual(name = "PTCL", labels = c("ALCL", "oPTCL"), values=c("#9b4ddb", "#37a695")) + theme_bw(base_size = 20) + xlab("") + stat_compare_means(comparisons = my_comparisons, size=4)+scale_x_discrete(labels=c("ALCL","oPTCL")) + theme(legend.position="none")
```


### 4.2. Graphical overview of ALCL VS AITL only for each cell type (Boxplot Supplementary Fig. 1)

```{r, warning = FALSE}

# create a dataframe for ggplot2 figures
d <- as.data.frame(cbind(sample=rownames(TME_estim_short),TME_estim_short)) %>% 
gather(key = "cell_type", value = "score", 2:25, na.rm = TRUE, factor_key=TRUE)
d$score <- as.numeric(d$score)
d_metadat <- cbind(cbind(patho=rep(metadata_short$patho,24),specific_patho=rep(metadata_short$specific_patho,24)),d)
d_metadat <- d_metadat[order(d_metadat$patho),]
d_metadat$sample <- factor(d_metadat$sample, levels = unique(d_metadat$sample))

# compute p-values between ALCL and AITL groups for the different cell types
p <- perform_wilcox_test(d, metadata_short)
# significative
p[which(p<0.0000001)] #1^10-7
pdf("boxplot_supp1.pdf", width = 10, height = 6)
type="T_cells_follicular_helper"
d_short <- d_metadat[which(d_metadat$cell_type==type),]
my_comparisons <- list( c("ALCL", "oPTCL"))
ggplot(d_short, aes(x=patho, y=score, fill=patho)) +  labs(title=type) + geom_boxplot(outlier.colour="darkgrey",outlier.size=2) + scale_fill_manual(labels = c("ALCL", "AITL"), values=c("#9b4ddb", "#96b0a1")) + theme_bw(base_size = 20) + xlab("") + stat_compare_means(comparisons = my_comparisons, size=4) + scale_x_discrete(labels=c("ALCL","AITL")) + theme(legend.position="none")
dev.off() 
```

